Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package \vskip15pt Supplementary Materials

### 2018-06-21 docs/MEE_submission/manuscript_MEE.tex
MS updates

\usepackage{amssymb,rotating,natbib,graphicx,fancyvrb}

\usepackage[parfill]{parskip} 
\usepackage[utf8]{inputenc}

@@

Although there are several tools that extract data from figures (e.g. \proglang{DataThief} \citep{DataThief}, \proglang{GraphClick} \citep{GraphClick}, \proglang{WebPlotDigitizer} \citep{WebPlotDigitizer}), these tools do not cater to needs of meta-analysis for four main reasons (here we focus on meta-analysis, although many  points apply to extraction for comparative analysis). First, although meta-analysis is an important tool in consolidating the data from multiple studies, many of the processes involved in data extraction are opaque and difficult to reproduce, making extending or replicating studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and aid in resolving the reproducibility crises seen in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, Parker2016}. Second, digitising programs do not allow the integration of metadata at the time of data extraction, such as experimental group or variable names, and sample sizes. This makes the downstream calculations laborious, as information has to be added later using different software. Third, existing programs do not import sets of images for the user to systematically work through. Instead they require the user to manually import images one by one, and export data into individual files, that need to be imported and edited using different software. Finally, digitising programs typically only provide the user with calibrated \textit{x,y} coordinates from imported figures, and do not differentiate between common plot types that are used to present data. Consequently a large amount of additional data manipulation is required, that is different across plots types. For example, data are frequently presented in plots with means and standard errors or confidence intervals (Figure \ref{fig:all_extract}A), from which the user wants a mean and SD for each group presented. From \textit{x,y} coordinates, users must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to SD, according to the type of error presented.

@@

As figures may have been copied from older, scanned publications, they may not be perfectly orientated. This makes calibration of the points in the figure problematic. \fct{metaDigitise} allows users to rotate the image (Figure \ref{fig:rotate}A,B). Furthermore, mean/error plots, box plots and histograms, may be presented with horizontal bars. \fct{metaDigitise} assumes that bars are vertical, but allows the user to flip the image to make the bars are vertical (Figure \ref{fig:rotate}C,D).

@@

\begin{figure}[!b] 
\centering
 \includegraphics[width=0.75\textwidth]{fig_rotate.pdf} 
 \caption{Figure rotation. A) and B) show how non-aligned images can be realigned through user defined rotation. C) and D) show how figures can be re-orientated so as to aid data input.}
\label{fig:rotate}
\end{figure}

>>>>

\usepackage{amssymb,rotating,natbib,graphicx,fancyvrb,booktabs,xcolor} 
\usepackage[parfill]{parskip} 
\usepackage[utf8]{inputenc}

@@

Although there are several tools that extract data from figures (e.g. \proglang{DataThief} \citep{DataThief}, \proglang{GraphClick} \citep{GraphClick}, \proglang{WebPlotDigitizer} \citep{WebPlotDigitizer}, see Table \ref{tab:comparison}), these tools do not cater to needs of meta-analysis for four main reasons (here we focus on meta-analysis, although many  points apply to extraction for comparative analysis). First, although meta-analysis is an important tool in consolidating the data from multiple studies, many of the processes involved in data extraction are opaque and difficult to reproduce, making extending or replicating studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and aid in resolving the reproducibility crises seen in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, Parker2016}. Second, digitising programs do not allow the integration of metadata at the time of data extraction, such as experimental group or variable names, and sample sizes. This makes the downstream calculations laborious, as information has to be added later using different software. Third, existing programs do not import sets of images for the user to systematically work through. Instead they require the user to manually import images one by one, and export data into individual files, that need to be imported and edited using different software. Finally, digitising programs typically only provide the user with calibrated \textit{x,y} coordinates from imported figures, and do not differentiate between common plot types that are used to present data. Consequently a large amount of additional data manipulation is required, that is different across plots types. For example, data are frequently presented in plots with means and standard errors or confidence intervals (Figure \ref{fig:all_extract}A), from which the user wants a mean and SD for each group presented. From \textit{x,y} coordinates, users must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to SD, according to the type of error presented.

@@

As figures may have been copied from older, scanned publications, they may not be perfectly orientated. This makes calibration of the points in the figure problematic. \fct{metaDigitise} allows users to rotate the image (Figure S2A,B). Furthermore, mean/error plots, box plots and histograms, may be presented with horizontal bars. \fct{metaDigitise} assumes that bars are vertical, but allows the user to flip the image to make the bars are vertical (Figure S2C,D).

@@

\clearpage

\section*{Tables}

\begin{table}[h!]
\centering
\rotatebox{90}{%
\begin{minipage}{1.35\textwidth}
\small
\begin{tabular}{lccccccc}
\hline
Function                  & metaDigitise & GraphClick$^1$ & DataThief$^2$  & DigitizeIt$^3$ & WebPlotDigitizer$^4$ & metagear$^5$ & digitize$^6$\\
\hline
Scatterplots              & \checkmark   & \checkmark & \checkmark & \checkmark & \checkmark     & \checkmark$^7$ & \checkmark \\
Mean/error plots      & \checkmark   & \checkmark & \checkmark & $\color{red}\times$   & $\color{red}\times$       & \checkmark$^7$ & $\color{red}\times$    \\
Boxplots                  & \checkmark   & $\color{red}\times$   & $\color{red}\times$   & $\color{red}\times$   & $\color{red}\times$       & $\color{red}\times$       & $\color{red}\times$    \\
Histograms                & \checkmark   & $\color{red}\times$   & $\color{red}\times$   & $\color{red}\times$   & \checkmark$^7$ & $\color{red}\times$       & $\color{red}\times$    \\
Graph rotation$^8$        & \checkmark   & \checkmark & \checkmark & \checkmark & \checkmark     & $\color{red}\times$       & $\color{red}\times$    \\
Grouped Data            & \checkmark   & \checkmark & $\color{red}\times$   & \checkmark & \checkmark     & $\color{red}\times$       & $\color{red}\times$    \\
Entry of metadata         & \checkmark   & $\color{red}\times$   & $\color{red}\times$   & $\color{red}\times$   & $\color{red}\times$       & $\color{red}\times$       & $\color{red}\times$    \\
Summarising data          & \checkmark   & $\color{red}\times$   & $\color{red}\times$   & $\color{red}\times$   & $\color{red}\times$       & $\color{red}\times$       & $\color{red}\times$    \\
Multiple image processing & \checkmark   & $\color{red}\times$   & $\color{red}\times$   & $\color{red}\times$   & $\color{red}\times$       & $\color{red}\times$       & $\color{red}\times$    \\
Reproducable$^9$  & \checkmark   & \checkmark & \checkmark & $\color{red}\times$   & \checkmark     & \checkmark       & $\color{red}\times$    \\
Automated point detection & $\color{red}\times$     & \checkmark &     $\color{red}\times$      & \checkmark & \checkmark     & \checkmark     & $\color{red}\times$    \\
Line extraction           & $\color{red}\times$     & \checkmark & \checkmark & \checkmark & \checkmark     & $\color{red}\times$       & $\color{red}\times$    \\
Zoom                      & $\color{red}\times$     & \checkmark & \checkmark & \checkmark & \checkmark     & $\color{red}\times$       & $\color{red}\times$    \\
Log axis                  & \checkmark   & \checkmark & \checkmark & \checkmark & \checkmark     & $\color{red}\times$       & $\color{red}\times$    \\
Dates                     & $\color{red}\times$   & $\color{red}\times$   & \checkmark & $\color{red}\times$   & \checkmark     & $\color{red}\times$       & $\color{red}\times$    \\
Asymmetric error bars     & $\color{red}\times$     & $\color{red}\times$   & \checkmark & $\color{red}\times$   & $\color{red}\times$       & $\color{red}\times$       & $\color{red}\times$    \\
Freeware                  & \checkmark$^{10}$ & \checkmark$^{11}$ & \checkmark$^{11}$ & $\color{red}\times$$^{11}$ & \checkmark$^{11}$ & \checkmark$^{10}$ & \checkmark$^{10}$\\
\hline

%http://datathief.org/DatathiefManual.pdf
%http://www.digitizeit.de/#Features
%https://automeris.io/WebPlotDigitizer/userManual.pdf

\end{tabular}
\\
$^1$ \citet{GraphClick}
$^2$ \citet{DataThief}
$^3$ \citet{DigitizeIt}
$^4$ \citet{WebPlotDigitizer}
$^5$ \citet{Lajeunesse2016}
$^6$ \citet{Poisot2011}
\\$^7$ Only automated, no manual extraction.
\\$^8$ Or handles rotated graphs. 
\\$^9$ Allows saving, re-plotting and editing of data extraction.
\\$^{10}$ \proglang{R} package.
\\$^{11}$ Standalone software.
\caption{\label{tab:comparison} Comparison of functionality between different digitisation softwares.}
\end{minipage}
}

\end{table}

### 2018-06-21 docs/MEE_submission/supplements.tex
MS updates

\usepackage{amssymb,rotating,natbib,graphicx,fancyvrb,booktabs}

\usepackage[parfill]{parskip} 
\usepackage[utf8]{inputenc}

@@

\fct{metaDigitise} then asks whether the figure needs to be rotated or flipped. This can be needed when box plots and mean and error plots are not orientated correctly. In some cases, older papers can give slightly off angled images which can be corrected by rotating. In this prompt the user has three options: \code{f} for ``Flip'', \code{r} for ``rotate'' or \code{c} for ``continue''.

@@

\section{Comparison with existing software}
\begin{table}[h!]
\centering
\rotatebox{90}{%
\begin{minipage}{1.35\textwidth}
\small
\begin{tabular}{lccccccc}
\hline
Function                  & metaDigitise & GraphClick$^1$ & DataThief$^2$  & DigitizeIt$^3$ & WebPlotDigitizer$^4$ & metagear$^5$ & digitize$^6$\\
\hline
Scatterplots              & \checkmark   & \checkmark & \checkmark & \checkmark & \checkmark     & \checkmark$^7$ & \checkmark \\
Mean/error plots      & \checkmark   & \checkmark & \checkmark & $\times$   & $\times$       & \checkmark$^7$ & $\times$    \\
Boxplots                  & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Histograms                & \checkmark   & $\times$   & $\times$   & $\times$   & \checkmark$^7$ & $\times$       & $\times$    \\
Graph rotation$^8$        & \checkmark   & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Groups                    & \checkmark   & \checkmark & $\times$   & \checkmark & \checkmark     & $\times$       & $\times$    \\
Entry of metadata         & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Summarising data          & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Multiple image processing & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Reproducable$^9$  & \checkmark   & \checkmark & \checkmark & $\times$   & \checkmark     & $\times$       & $\times$    \\
Automated point detection & $\times$     & \checkmark &     $\times$      & \checkmark & \checkmark     & \checkmark     & $\times$    \\
Line extraction           & $\times$     & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Zoom                      & $\times$     & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Log axis                  & $\times$   & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Dates                     & $\times$   & $\times$   & \checkmark & $\times$   & \checkmark     & $\times$       & $\times$    \\
Asymmetric error bars     & $\times$     & $\times$   & \checkmark & $\times$   & $\times$       & $\times$       & $\times$    \\
Freeware                  & \checkmark$^{10}$ & \checkmark$^{11}$ & \checkmark$^{11}$ & $\times$$^{11}$ & \checkmark$^{11}$ & \checkmark$^{10}$ & \checkmark$^{10}$\\
\hline

%http://datathief.org/DatathiefManual.pdf
%http://www.digitizeit.de/#Features
%https://automeris.io/WebPlotDigitizer/userManual.pdf

\end{tabular}
\\
$^1$ \citet{GraphClick}
$^2$ \citet{DataThief}
$^3$ \citet{DigitizeIt}
$^4$ \citet{WebPlotDigitizer}
$^5$ \citet{Lajeunesse2016}
$^6$ \citet{Poisot2011}
\\$^7$ Only automated, no manual extraction.
\\$^8$ Or handles rotated graphs. 
\\$^9$ Allows saving, re-plotting and editing of data extraction.
\\$^{10}$ \proglang{R} package.
\\$^{11}$ Standalone software.
\caption{\label{tab:comparison} Comparison of functionality between different digitisation softwares.}
\end{minipage}
}

\end{table}

>>>>

\usepackage{amssymb,rotating,natbib,graphicx,fancyvrb,booktabs} 
\usepackage[parfill]{parskip} 
\usepackage[utf8]{inputenc}

@@

\fct{metaDigitise} then asks whether the figure needs to be rotated or flipped. This can be needed when box plots and mean and error plots are not orientated correctly (as shown in Figure \ref{fig:rotate}). In some cases, older papers can give slightly off angled images which can be corrected by rotating. In this prompt the user has three options: \code{f} for ``Flip'', \code{r} for ``rotate'' or \code{c} for ``continue''.

\begin{figure}[!b] 
\centering
 \includegraphics[width=0.75\textwidth]{fig_rotate.pdf} 
 \caption{Figure rotation. A) and B) show how non-aligned images can be realigned through user defined rotation. C) and D) show how figures can be re-orientated so as to aid data input.}
\label{fig:rotate}
\end{figure}

@@
